---
title: "An exploration of Alchohol Consumption and Average Grades"
author: "Jazmine(Junzhi) Sun and Heriberto (Heri) Lopez"
output:
  pdf_document:
    fig_height: 3
    fig_width: 4.5
  html_document: default
  word_document: default
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
editor_options: 
  chunk_output_type: console
---
```{r, include = FALSE}
library(mosaic)
library(ggplot2)
library(utils)
library(tidyverse)
library(readr) 
library(gridExtra)
library(glmnet)

#Our two Datasets from Kaggle
Student_por <- read_csv("~/Stats 272 S20/Project/HeriandJazmine/STAT-272-Project/DataSets/student_por.csv")
Student_mat <- read_csv("~/Stats 272 S20/Project/HeriandJazmine/STAT-272-Project/DataSets/student_mat.csv")

# Average of g1, g2, and g3 for every student, This wil serves as our response variable in our model 
Student_mat <- Student_mat %>%
  mutate(AvgGrade = (g1 + g2 + g3)/3,
         sex_ind = ifelse(sex == "F", 1, 0)) 
Student_por <- Student_por %>%
  mutate(AvgGrade = (g1 + g2 + g3)/3, 
         sex_ind = ifelse(sex == "F", 1, 0)) 

```
\begin{center}
  **Introduction**
\end{center} 
Colleges, Universities, and many schools across the world have implemented policies to tackle issues regarding alcohol use on their campuses. Many policies come with an effort to influence students to have control over their drinking habits and enjoy activities without feeling the need to feel inebriated. While schools seem to focus on the social aspects and harmful effects on individuals outside of academic life researchers have developed a keen interest on the extent of the affect that alcohol has on the academic performance of students. Research on the effects of alcohol consumption has been widely studied. Large areas of studies have focused on the social factors that may lead to binge drinking as students. A study published in 2014, “Just a First-Year Thing? The Relations between Drinking During Orientation Week and Subsequent Academic Year Drinking Across Class Years”, gathered undergraduate participants from the University of Otago, New Zealand and collected survey data on the student participants for 13 days through daily diaries which included personal amounts of alcohol consumption. The study focused mainly on patterns for event specific drinking such as Orientation week. (Riordan, Benjamin, et. all). The results suggest that event-specific drinking increases the amount of alcohol consumption across all class years. Similarly, other studies have looked at predictors surrounding alcohol and the association with poor academic achievement through survey and academic data of students (Walid, et al). But to what extent does alcohol influence academic performance. Of interest when looking at the effects of alcohol are potentially other factors that can be attributed to a change in academic performance. Joung Yeoub and Austin Smith in there research article, “Legal access to alcohol and academic performance: Who is affected?”, published in the Economics of Education review use survey data to compare the average drinks students of legal age compare to those of younger ages results were similar to other studies where peers of legal drinking age showed a significant increase when obtained the legal drinking age compared to their younger peers. Furthermore, to look at the affect more accurately the researchers created a model to predict the impact of legal access on alcohol consumption by creating an indicator variable for those of legal age and able to have swift access to drinks during the semester, an individual effect variable that accounted for an unobserved time variant that may have an effect on grade, a variable for semester characteristic variants, and a variable on individual effects of academic performance. 
For this project, we will be implementing a similar approach on analyzing student alcohol consumption from the ‘Student Alcohol consumption Dataset in Kaggle and focusing on other predictors as well that are strong indicators in predicting the final grade of the students in this study. The dataset includes grades from Portuguese students in secondary school from a respective mathematics course and a language course. There was no missing data present in our dataset and data was collected through surveys. The grading system in Portugal secondary schools are graded differently than in the United States grades and are calculated between a 0 through 20 point scale (A = 20 - 18, B = 17 - 16, C = 15 -14, D = 13-12, E (Sufficient) = 11-10). All statistical computing was calcualted using R software. 

\begin{center}
Materials and Methods
\end{center}

While the model produced by researchers Joung Yeoub and Austin Smith highlighted the possible relationships between legal age of drinking and individual characteristics the study focused on older students and mainly those residing in the United States. Due to our dataset, we were interested in looking at affects to the average final grade for the secondary school students in our dataset. We began by calculating a final average over the three periods of the academic school year that the students were graded for in the mathematics course and the language course. We calculated the average final grade to compare the affect of alcohol consumption and other predictors over the whole academic year.

```{r, include = FALSE}
# Histograms of average grade 
plot1a <- ggplot(data = Student_mat, mapping = aes(x = AvgGrade)) +
  geom_histogram(bins = 20, fill = "Blue", color = "Black") + 
  labs(x ="Average Grade")
  

# In Student_mat
plot1b <- ggplot(data = Student_por, mapping = aes(x = AvgGrade)) +
  geom_histogram(bins = 20, fill = "Red", color = "Black") + 
  labs(x = "Average Grade") # In Student_por  
# Summary statistics
favstats(~ AvgGrade, data = Student_mat) # In Student_mat
favstats(~ AvgGrade, data = Student_por) # In Student_por
```

```{r, message = FALSE, echo = TRUE, results = 'hide'} 
grid.arrange(plot1a,plot1b, nrow = 2) 
```

**Figure 1.** A distribution of average grades for the mathematics course (Blue) and the language course (Red). The mathematics course seems to have normal distribution while the distribution of the Portuguese language class seems to be slightly skewed.

```{r, message = FALSE, echo = TRUE, include = FALSE}
# Figure 1a
favstats(~ AvgGrade, data = Student_mat) # In Student_mat
favstats(~ AvgGrade, data = Student_por)
#Student_por  
649 - 395 

ggplot(data = Student_mat, mapping = aes(x = AvgGrade, color = sex, fill = sex)) + 
geom_bar() + 
  facet_wrap(~sex, ncol = 2)

ggplot(data = Student_mat, mapping = aes(x = sex, fill = sex)) + 
  geom_bar() + 
  labs(x = "sex of students in the mathematics course") 

ggplot(data = Student_por, mapping = aes(x = sex, fill = sex)) + 
  geom_bar() + 
  labs(x = "sex of students in the language course")

Student_mat %>% 
  group_by(sex) %>% 
  count()


Student_por %>% 
  group_by(sex) %>% 
  count() 

Student_mat %>% 
  group_by(school, sex) %>% 
  count()

```

 Further summary statistics show that mean grade for both courses were relatively close to each other (math mean = 10.33, language mean = 11.67). There were 395 students taking a mathematics course and 649 taking a Portuguese language course. 208 of the students taking a mathematics course were female and 187 were male. In the language course 383 students were female and 266 students were male. The data was gathered from two schools the Gabriel Pereira (GP) and the Mousinho da Silveira (MS) schools in Portugal. In both courses there was 382 students who were taking both the mathematics course and language course. 

To isolate the effect of alcohol consumption, we explored relationships between other potential predictors for average of the students. We looked at predictors that have been associated with grades in the past such as the highest education achieved by both of the student's parents, the number of absences stundets have and others. We compared them with each other predictors to see if there were interactions between variables as well. We found that for both mother and father’s education there were less than 10 students whose parents had no education at all. Because of these initial graphs of parents education showed a larger affect coming from the 0 group. 

```{r, echo = TRUE, include = FALSE}
Student_mat %>% 
  group_by(medu, sex) %>% 
  count(sex) 

Student_por %>% 
  group_by(medu, sex) %>% 
  count() 

ggplot(data = Student_mat, mapping = aes(x = as.factor(medu), fill = sex)) + 
  geom_bar(position = "fill") 

ggplot(data = Student_por, mapping = aes(x = as.factor(medu), fill = sex)) + 
  geom_bar(position = "fill") 

Student_mat<- Student_mat %>% 
  mutate(medu = ifelse(medu == 0, NA, medu), 
         fedu = ifelse(fedu == 0, NA, fedu)) %>%
  filter(!is.na(medu)) %>% 
  filter(!is.na(fedu)) 

Student_por <- Student_por %>% 
  mutate(medu = ifelse(medu == 0, NA, medu), 
         fedu = ifelse(fedu == 0, NA, fedu)) %>%
  filter(!is.na(medu)) %>% 
  filter(!is.na(fedu)) 

plot3a <- ggplot(data = Student_mat, mapping = aes(x = as.factor(medu), y = AvgGrade, )) +
  geom_boxplot(fill = "Blue") + 
  labs(x = "Mother's Education")# In Student_mat
plot3b <- ggplot(data = Student_por, mapping = aes(x = as.factor(medu), y = AvgGrade)) +
  geom_boxplot(fill = "Red") + 
  labs(x = "Mother's Education")# In Student_por 
plot3c <- ggplot(data = Student_mat, mapping = aes(x = as.factor(fedu), y = AvgGrade, )) +
  geom_boxplot(fill = "Blue") + 
  labs(x = "Father's Education")
plot3d <- ggplot(data = Student_por, mapping = aes(x = as.factor(fedu), y = AvgGrade, )) +
  geom_boxplot(fill = "red") + 
  labs(x = " Father's Education")

```

```{r, include = FALSE}
grid.arrange(plot3a,plot3b, ncol = 2)
# Summary statistics
favstats(~ medu, data = Student_mat) # In Student_mat
favstats(~ medu, data = Student_por) # In Student_por
``` 

```{r}
grid.arrange(plot3a,plot3b, plot3c, plot3d, ncol = 4)
```

**Figure 3.** The boxplots show the average grade for students with different levels of Mother's education. This plot include students whose parents have a zero education level.  

```{r, include = FALSE}
Student_mat %>% group_by(sex) %>% 
  summarise(absences = sum(absences),
            n = count(absences))
Student_por %>% 
  group_by(sex) %>% 
  summarise(absences = sum(absences),
            n = count(absences))

favstats(~ absences, data = Student_mat) 
favstats(~absences, data = Student_por) 

ggplot(data = Student_mat, mapping = aes(x = absences)) + 
  geom_histogram() + 
  facet_wrap(~sex, ncol = 2) 

Student_mat <- Student_mat %>% 
  mutate(
    absences = ifelse(absences > 4, "high", "low") 
  )  
Student_por <- Student_por %>% 
  mutate(
    absences = ifelse(absences > 2, "high", "low") 
  )  

ggplot(data = Student_mat, mapping = aes(x = absences, y = AvgGrade, fill = absences)) + 
  geom_boxplot() + 
    facet_wrap(~sex, nrow = 2)

ggplot(data = Student_por, mapping = aes(x = absences, y = AvgGrade, fill = absences)) + 
  geom_boxplot() + 
  facet_wrap(~sex, nrow = 2) pp


  
```

Overall, as the level of education in the parents increases the average grade is also increasing. When looking at absences the average number of absences was fairly similar in both courses with the mathematics course having a slighlty higher average (mean = 5.75, sd = 8) than the language course (mean = 3.7, sd = 4.7) and females having more absences than their male counterparts. When comparing the grades of students with high absences there seemed to be no difference between the medians. Furthermore, the distribution of average grade in both the math and language course seemed to have a normal distrbution even when looking at the differences between sex. Before moving on to our main predictor, we explored the relationship between age. The average age in both courses is approximately 16 years of age in both courses. We noted that the legal drinking age for alcohol in Portugal is 18 years of age. In both courses more than half of the students indicated that they were underage. We explored this with average age obtained a slight difference between the medians. When looking at the differences between sex the underage group for males tended to perform better than the legal group. There were similar difference seen in the language course. 


```{r, include = FALSE}

plot4c<- ggplot(data = Student_mat, mapping = aes(x = AvgGrade)) + 
  geom_histogram() + 
  facet_wrap(~sex, ncol = 2) + 
  labs(x = "Mean Grade for students in Mathematics Course")

plot4d<-ggplot(data = Student_por, mapping = aes(x = AvgGrade)) + 
  geom_histogram() + 
  facet_wrap(~sex, ncol = 2) + 
  labs(x = "Mean Grade for students in Language Course") 

```

```{r, message = FALSE, echo = TRUE, include = FALSE}
grid.arrange(plot4c, plot4d, nrow = 2) 

#**Figure 4.** Histogram of Average grade by sex (F = Female, M = Male)
#Our main predictor on alcohol consumption focused on the amount of workday drinking and weekend drinking. We wanted to make sure that there was not an intereaction variable present and so we explored the realtionship between other predictors with the two alcohol predictors. When looking at sex we found that the distribution follows a similar pattern for both male and femlae with more females listing themselves in the low level consumption group, and less than 25 people listing themselves in the high level of consumption. This trend is similar in both courses. 
```

```{r, include = FALSE}

#**Figure 4.** Histogram of Average grade by sex (F = Female, M = Male)

#There is no missing variable of explanatory variable in both two datasets. By the box plots, the median average grade for students in math class do not have huge difference when the weekend alcohol consumption level are 1, 2, and 3. When the eekend alcohol consumption are 4 and 5, which levels are high and very high, the median average grade decreases. The mean level of work day alcohol consumption is 1.48. The median average grade for students in Potuguese language also do not have huge difference when the workday alcohol consumption level are 1, 2, and 3. The mean level of work day alcohol consumption is  1.50. 

```

```{r, include = FALSE}

plot4c<- ggplot(data = Student_mat, mapping = aes(x = AvgGrade)) + 
  geom_histogram() + 
  facet_wrap(~sex, ncol = 2) + 
  labs(x = "Mean Grade for students in Mathematics Course")

plot4d<-ggplot(data = Student_por, mapping = aes(x = AvgGrade)) + 
  geom_histogram() + 
  facet_wrap(~sex, ncol = 2) + 
  labs(x = "Mean Grade for students in Language Course") 

```

```{r, message = FALSE, echo = TRUE, include = FALSE}
grid.arrange(plot4c, plot4d, nrow = 2)

#*Figure 4.** Histogram of Average grade by sex (F = Female, M = Male)

#
```

```{r, include = FALSE}

#**Figure 4.** Histogram of Average grade by sex (F = Female, M = Male)

#There is no missing variable of explanatory variable in both two datasets. By the box plots, the median average grade for students in math class do not have huge difference when the weekend alcohol consumption level are 1, 2, and 3. When the eekend alcohol consumption are 4 and 5, which levels are high and very high, the median average grade decreases. The mean level of work day alcohol consumption is 1.48. The median average grade for students in Potuguese language also do not have huge difference when the workday alcohol consumption level are 1, 2, and 3. The mean level of work day alcohol consumption is  1.50. 

```

```{r, include = FALSE}
Student_mat
favstats(~age, data = Student_mat) 
favstats(~age, data = Student_por) 

Student_mat <- Student_mat %>% 
  mutate(legal = ifelse(age >= 18, "legal", "underage"))  
Student_por<- Student_por %>% 
  mutate(legal = ifelse(age >= 18, "legal", "underage")) 

Student_mat %>% 
  count(legal)
Student_por %>% 
  count(legal)

Student_mat %>% 
  group_by(sex) %>% 
  count(legal) 
Student_por %>% 
  group_by(sex) %>% 
  count(legal) 

plot5a<- ggplot(data = Student_mat, mapping = aes(x = legal, y = AvgGrade)) + 
  geom_boxplot() + 
  facet_wrap(~sex, ncol = 2)
 

plot5b<- ggplot(data = Student_por, mapping = aes(x = legal, y = AvgGrade)) + 
  geom_boxplot() + 
  facet_wrap(~sex, ncol = 2)

```

```{r, include = FALSE}

ggplot(data = Student_mat, mapping = aes(x = as.factor(dalc), y = AvgGrade)) +
  geom_boxplot()
  facet_wrap(~sex, ncol = 2) 
  
ggplot(data = Student_, mapping = aes(x = as.factor(dalc), y = AvgGrade)) +
  geom_boxplot()
  facet_wrap(~sex, ncol = 2)  
    
ggplot(data = Student_mat, mapping = aes(x = dalc)) + 
  geom_bar() + 
  facet_wrap(~sex, ncol = 2)  


ggplot(data = Student_por, mapping = aes(x = dalc)) + 
  geom_bar() + 
  facet_wrap(~sex, ncol = 2)  

Student_mat <- Student_mat %>% 
  mutate(workalc = ifelse(dalc %in% c("1,", "2"), "low", "high"), 
         wkndalc = ifelse(walc %in% c("1,", "2"), "low", "high"))

Student_por <- Student_por %>% 
  mutate(workalc = ifelse(dalc %in% c("1,", "2"), "low", "high"), 
         wkndalc = ifelse(walc %in% c("1,", "2"), "low", "high")) 

ggplot(data = Student_mat, mapping = aes(x = workalc, y = AvgGrade)) + 
  geom_boxplot() + 
  facet_wrap(~sex, ncol = 2) 

ggplot(data = Student_mat, mapping = aes(x = wkndalc, y = AvgGrade)) + 
  geom_boxplot() +
  facet_wrap(~sex, ncol = 2)

ggplot(data = Student_por, mapping = aes(x = workalc, y = AvgGrade)) + 
  geom_boxplot() + 
  facet_wrap(~sex, ncol = 2) 

ggplot(data = Student_por, mapping = aes(x = wkndalc, y = AvgGrade)) + 
  geom_boxplot() +
  facet_wrap(~sex, ncol = 2)  

################legal with wrk

ggplot(data = Student_mat, mapping = aes(x = workalc, y = AvgGrade)) + 
  geom_boxplot() + 
  facet_wrap(~legal, ncol = 2) 

ggplot(data = Student_mat, mapping = aes(x = wkndalc, y = AvgGrade)) + 
  geom_boxplot() +
  facet_wrap(~legal, ncol = 2)

ggplot(data = Student_por, mapping = aes(x = workalc, y = AvgGrade)) + 
  geom_boxplot() + 
  facet_wrap(~legal, ncol = 2) 

ggplot(data = Student_por, mapping = aes(x = wkndalc, y = AvgGrade)) + 
  geom_boxplot() +
  facet_wrap(~legal, ncol = 2) 

##############Absences 

ggplot(data = Student_mat, mapping = aes(x = workalc, y = AvgGrade)) + 
  geom_boxplot() + 
  facet_wrap(~absences, ncol = 2) 

ggplot(data = Student_mat, mapping = aes(x = wkndalc, y = AvgGrade)) + 
  geom_boxplot() +
  facet_wrap(~absences, ncol = 2)

ggplot(data = Student_por, mapping = aes(x = workalc, y = AvgGrade)) + 
  geom_boxplot() + 
  facet_wrap(~absences, ncol = 2) 

ggplot(data = Student_por, mapping = aes(x = wkndalc, y = AvgGrade)) + 
  geom_boxplot() +
  facet_wrap(~absences, ncol = 2)  


```

```{r}
grid.arrange(plot5a, plot5b, ncol = 2)
```

**Figure 4.** Boxplot of legal and underage drinking ages and average grade. 

We finally looked at our main predictor alcohol consumption. There were two alcohol predictors in our data workday and weekend alcohol consumptions. We compared these to each of the other predictors to account for any associations between variables, but first began by looking at their relationship with the average grade. 

```{r} 

```


We also want to look at the effect of mother's education on average grade. Figure 3. shows the education level of the Mother's education and the average grade. In both cases, the lowest average grade median comes occurs when the mother has only completed primary school. 

For the effect of father's eductaion on average grade. Figure 4. shows boxplots for the mathematics course (Blue) and the language class (Red). The median average grade seems to be lower again when the father has only completed primary education


\begin{center}
Initial Models
\end{center}

According to the previous EDA, we notice that there exists a negative relationship between average grade and the alcohol consumption level. So, firstly, we choose alcohol consumption as the potential explanatory variable that has an affect on the average grade.

Because there are two kinds of alcohol consumption in the datasets, we will build two models based on different alcohol consumption. We build *Model1*, *Model2*, and *Model3* in the dataset for students in math class. In Model1, we have workday alcohol consumption level as the explanatory variable, and we try to find the relationship between workday alcohol consumption and the average grade. We have the linear regression $\hat{AvgGrade} = 11.1250 - 0.3009 dalc$. $\hat\beta_1 = -0.3009$: The average grade decreases 0.3009, on average, as 1 level increases in workday alcohol consumption. In Model2, we take weekend alcohol consumption level as an explanatory variable in this model. We have linear regression $\hat{AvgGrade} = 11.2582 - 0.2527 walc$. $\hat\beta_1 = -0.2527$: The average grade decreases 0.2527, on average, as 1 level increases in weekend alcohol consumption. Then, we can build the model which takes two kinds of alcohol consumption as explanatory variables in Model 3. We have linear regression $\hat{AvgGrade} = 11.3086 - 0.1108 dalc - 0.2030 walc$. $\hat\beta_1 = -0.1108$: The average grade decreases 0.1108, on average, as 1 level increases in workday alcohol consumption after controlling the level of weekend alcohol consumption as a constant. $\hat\beta_2 = -0.2030$: The average grade decreases 0.2030, on average, as 1 level increases in weekend alcohol consumption after controlling the level of workday alcohol consumption as a constant. 

Then, we build *Model1a*, *Model2a*, and *Model3a* in the dataset for students in the Portuguese language class. We use the similar method that we use to build models in Student_mat. We can notice that there exists a similar relationship between the average grades and alcohol consumption. In Model1a, we have workday alcohol consumption level as the explanatory variable.  We have the linear regression $\hat{AvgGrade} = 12.5722 - 0.6305 dalc$. $\hat\beta_1 = -0.6305$: The average grade decreases 0.5305, on average, as 1 level increases in workday alcohol consumption. In Model2a, we take weekend alcohol consumption level as an explanatory variable in this model. We have linear regression $\hat{AvgGrade} = 12.49994 - 0.38364 walc$. $\hat\beta_1 = -0.38364$: The average grade decreases 0.38364, on average, as 1 level increases in weekend alcohol consumption. Then, we can build the model which takes two kinds of alcohol consumption as explanatory variables in Model 3a. We have linear regression $\hat{AvgGrade} = 12.7386 - 0.4872 dalc -0.1673walc$. $\hat\beta_1 = -0.4872$: The average grade decreases 0.4872, on average, as 1 level increases in workday alcohol consumption after controlling the level of weekend alcohol consumption as a constant. $\hat\beta_2 = -0.1673$: The average grade decreases 0.1673, on average, as 1 level increases in weekend alcohol consumption after controlling the level of workday alcohol consumption as a constant.

Next, we want to look the effects of mather's education and father's education on the average grades. In Model4, Model5, and Model6, we use other potential explanatory variables to build the models in Student_mat. We have mother’s education or father’s education as the predictor, and we contain both two predictors in Model6. In Model4, the explanatory variable is mother’s education. We have $\hat{AvgGrade} = 8.5972 + 0.7573 medu$. $\hat\beta_1 = 0.7573$: The average grade increases 0.7573, on average, as 1 level of mother’s education increases.  In Model5, the explanatory variable is father’s education. We have $\hat{AvgGrade} = 9.1730 + 0.5974 fedu$. $\hat\beta_1 = 0.5974$: The average grade increases 0.5974, on average, as 1 level increases in father’s education. In Model6, we have both mother’s education and father’s education as the explanatory variables. We have a linear regression line $\hat{AvgGrade} = 8.4335 + 0.6332 medu + 0.2003 fedu$. $\hat{beta_1} = 0.6332$: The average grade increases 0.6332, on average, as 1 level increases in mother’s education after controlling father’s education level as a constant. $\hat\beta_2 = 0.2003$: The average grade increases 0.2003, on average, as 1 level increases in father’s education after controlling mother’s education level as a constant. 

In Model4a, Model5a, and Model6a, we have mother’s education or father’s education as the predictor in Student_por. In Model4a, the explanatory variable is mother’s education. We have $\hat{AvgGrade} = 9.95521 + 0.66405 medu$. $\hat\beta_1 = 0.66405$: The average grade increases 0.0.66405, on average, as 1 level of mother’s education increases.  In Model5a, the explanatory variable is father’s education. We have $\hat{AvgGrade} = 10.27083 + 0.58710 fedu$. $\hat\beta_1 = 0.58710$: The average grade increases 0.58710 on average, as 1 level increases in father’s education. In Model6, we have both mother’s education and father’s education as the explanatory variables. We have a linear regression line $\hat{AvgGrade} = 9.7752+ 0.5088medu + 0.2473 fedu$. $\hat\beta_1 = 0.5088$: The average grade increases 0.5088, on average, as 1 level increases in mother’s education after controlling father’s education level as a constant. $\hat\beta_2 = 0.0.2473$: The average grade increases 0.2473on average, as 1 level increases in father’s education after controlling mother’s education level as a constant.


\begin{center}
**Future Plan**
\end{center}
In the future plan, we plan to add interaction terms in our model. For example, we could add absence as predictors in the models for the relationship between average grade and alcohol consumption and then add the interaction between alcohol consumption and absence. Because alcohol consumption may cause the absence and the absence could be related with higher or lower average grades. In order to build larger models in the future, we plan to find the correlation between predictors. We need to check whether the predictors are highly correlated with each other. We also can find if there are influential points in our models. If there exists an influential point in the model, we can move the influential points from the model and build a new one to find the relationship between response variable and explanatory variable(s) and do prediction. 

We also could include the guardian in our models. We could add interaction between mother's education or father’s education and the guardian. We could check whether the relationship between average grade and  parents’ education level can be affected by the guardian. Because $69\%$ students choose mother as the guardian in Student_mat and $70\%$ students choose mother as the guardian in Student_por. In this case, we guess that both guardian and parents’ education level can havel effect on the average grades of students. 


\begin{center} 
**Appendix** 
\end{center}

```{r, include = FALSE}
# Histograms of average grade 
plot1a <- ggplot(data = Student_mat, mapping = aes(x = AvgGrade)) +
  geom_histogram(bins = 20, fill = "Blue", color = "Black") # In Student_mat
plot1b <- ggplot(data = Student_por, mapping = aes(x = AvgGrade)) +
  geom_histogram(bins = 20, fill = "Red", color = "Black") # In Student_por  

# Summary statistics
favstats(~ AvgGrade, data = Student_mat) # In Student_mat
favstats(~ AvgGrade, data = Student_por) # In Student_por
```

```{r, message = FALSE, echo = TRUE} 
grid.arrange(plot1a,plot1b, ncol = 2)
```
**Figure 1.**   
```{r, message = FALSE, echo = TRUE, include = FALSE}
# Figure 1a
favstats(~ AvgGrade, data = Student_mat) # In Student_mat
favstats(~ AvgGrade, data = Student_por) #Student_por
```
**Figure1a** Summary statistics of average grade for the students in both classes.  

```{r, echo = TRUE, include = FALSE}
# Box plot of "workday alcohol consumption"
plot2a <- ggplot(data = Student_mat, mapping = aes(x = as.factor(dalc), y = AvgGrade, )) +
  geom_boxplot(fill = "Blue") + 
  coord_flip()# In Student_mat
plot2b <- ggplot(data = Student_por, mapping = aes(x = as.factor(dalc), y = AvgGrade)) +
  geom_boxplot(fill = "Red") + 
  coord_flip()# In Student_por

grid.arrange(plot2a,plot2b, ncol = 2)

# Summary statistics
favstats(~ dalc, data = Student_mat) # In Student_mat
favstats(~ dalc, data = Student_por) # In Student_por
```

```{r, message = FALSE, echo = TRUE, include = FALSE}
grid.arrange(plot2a,plot2b, ncol = 2) 

favstats(~ dalc, data = Student_mat) # In Student_mat
favstats(~ dalc, data = Student_por) # In Student_por
```
**Figure 2.**

```{r, echo = TRUE, include = FALSE}
# Box plot of "mather's education" 
plot3a <- ggplot(data = Student_mat, mapping = aes(x = as.factor(medu), y = AvgGrade, )) +
  geom_boxplot(fill = "Blue")# In Student_mat
plot3b <- ggplot(data = Student_por, mapping = aes(x = as.factor(medu), y = AvgGrade)) +
  geom_boxplot(fill = "Red")# In Student_por 
```

```{r}
grid.arrange(plot3a,plot3b, ncol = 2)
# Summary statistics
favstats(~ medu, data = Student_mat) # In Student_mat
favstats(~ medu, data = Student_por) # In Student_por
```

**Figure 3.**

```{r, echo = TRUE}
# Box plot of "father's education"
plot4a <- ggplot(data = Student_mat, mapping = aes(x = as.factor(fedu), y = AvgGrade, )) +
  geom_boxplot(fill = "Blue") + 
  coord_flip()# In Student_mat
plot4b <- ggplot(data = Student_por, mapping = aes(x = as.factor(fedu), y = AvgGrade)) +
  geom_boxplot(fill = "Red") + 
  coord_flip()# In Student_por
```

```{r}
grid.arrange(plot4a,plot4b, ncol = 2)

# Summary statistics
favstats(~ fedu, data = Student_mat) # In Student_mat
favstats(~ fedu, data = Student_por) # In Student_por
```
**Figure 4.** 
```{r, include = FALSE} 
Student_mat <-Student_mat %>% 
  mutate(sex_ind = ifelse(sex == "F", 1,0))

Student_por <- Student_mat %>% 
  mutate(sex_ind = ifelse(sex == "F", 1, 0))
```


```{r, echo = TRUE}
# "workday alcohol consumption" and "weekend alcohol consumption" as explanatory variables

# In Student_mat
Model1 <- lm(AvgGrade ~ dalc, data = Student_mat)
summary(Model1)
Model2 <- lm(AvgGrade ~ walc, data = Student_mat)
summary(Model2)
Model3 <- lm(AvgGrade ~ dalc + walc, data = Student_mat)
summary(Model3)
```
**Model1, Model2, and Model3**

```{r,echo = TRUE}
# In Student_por
Model1a <- lm(AvgGrade ~ dalc, data = Student_por)
summary(Model1a)
Model2a <- lm(AvgGrade ~ walc, data = Student_por)
summary(Model2a)
Model3a <- lm(AvgGrade ~ dalc + walc, data = Student_por)
summary(Model3a)
```
**Model1a, Model2a, and Model3a**

```{r, echo = TRUE}
# "mather's education" and "father's education" as explanatory variables

# In Student_mat
Model4 <- lm(AvgGrade ~ medu, data = Student_mat)
summary(Model4)
Model5 <- lm(AvgGrade ~ fedu, data = Student_mat)
summary(Model5)
Model6 <- lm(AvgGrade ~ medu + fedu, data = Student_mat)
summary(Model6)
```
**Model4, Model5, and Model6**
```{r,echo = TRUE}
# In Student_por
Model4a <- lm(AvgGrade ~ medu, data = Student_por)
summary(Model4a)
Model5a <- lm(AvgGrade ~ fedu, data = Student_por)
summary(Model5a)
Model6a <- lm(AvgGrade ~ medu + fedu, data = Student_por)
summary(Model6a)
```
**Model4a, Model5a, and Model6a**

```{r, echo = FALSE}
# Adding absences as the predictor
Model7 <- lm(AvgGrade ~ absences, data = Student_mat)
summary(Model7)

# Interaction between workday alcohol consumption and the absences
Model7a <- lm(AvgGrade ~ dalc + absences + dalc: absences, data = Student_mat)
summary(Model7a) # Not significant
Model7b <- lm(AvgGrade ~ walc + absences + walc: absences, data = Student_mat)
summary(Model7b) # Not significant

Model7c <- lm(AvgGrade ~ dalc + absences + dalc: absences, data = Student_por)
summary(Model7c)
Model7d <- lm(AvgGrade ~ walc + absences + walc: absences, data = Student_por)
summary(Model7d)
```

## Selecting Models !!!! Choose this one
```{r, include = FALSE}
Model0 <- lm(AvgGrade ~ 1, data = Student_mat)
summary(Model0)
Model1_initial <- lm(AvgGrade ~ dalc + walc + absences + fedu + medu + sex, data = Student_mat)
step(Model0, direction = "both", scope = formula(Model1_initial))
```
By selecting model, we can have the potential predictors are mother's education level, gender of the students, and weekend alcohol consumption in the dataset for students in math class. Then, we build the model contains these three variables. 
```{r, include = FALSE}
Model8 <- lm(AvgGrade ~ medu + sex + walc, data = Student_mat)
summary(Model8)
```

We add an interaction between gender and weekend alcohol consumption, in order to check whether the effect of workday alcohol consumption on the average grade depends on the leads of gender. 
```{r, include = FALSE}
# Adding the interaction between gender and workday alcohol consumption 
Model8_inter <- lm(AvgGrade ~ medu + sex + walc + sex:walc, data = Student_mat)
summary(Model8_inter)

anova(Model8, Model8_inter)
```
In Model8 with interaction, we have the euqation $\hat{AvgGrade} = 7.7573 + 0.7837 medu + 2.8489 sexM + 0.2268 walc - 0.9031 sexM:walc$. 

$\hat\beta_0 = 7.7573$: The expected average grade is $7.7573$ when female students with 0 in mother's education level and 0 in weekend alcohol consumption. 

$\hat\beta_1 = 0.7837$: The expected average grade increases $0.7837$, on average, as 1 levle increase in mother's education after controlling the gender of students and the weekend alcohol consumption. 

When the gender is female, we have the equation $\hat{AvgGrade} = 7.7573 + 0.7837 medu + 0.2268 walc$: When students is female, the expected average grade increases $0.2268$, on average, as 1 level increases in weekend alcohol consumption after controlling the level of mother's education. [!!!! Interesting results !!!]

When the gender is male, we have the equation $\hat{AvgGrade} = (7.7573 + 2.8489) + 0.7837 medu + (0.2268 - 0.9031) walc$: When students is male, the expected average grade decreases $0.6763$, on average, as 1 level increases in weekend alcohol consumption after controlling the level of mother's education. 

We have statistically significant evidence (t = -3.035, p = 0.00257) that the interaction between gender and weekedn alcohol consumption have effect on the response variable. Also, by anova table, we can say that Model8 with interaction has improvement (p = 0.002571) on Model8 without interaction.

### Residual Plot
```{r, echo = FALSE}
par(mfrow = c(2, 2), mar = c(2, 2, 2, 2)) 
plot(Model8_inter, which = c(1, 2, 3, 5))
par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

### Confidence Interval 
```{r, echo = FALSE}
confint(Model8_inter)
```
*medu*: we are $95\%$ confident that as 1 level increases in mother's education, the average grade increases between $0.453$ and $1.115$ after controlling the gender of students and weekend alcohol consumption. 
*walc*: we are $95\%$ confident that as 1 level increases in weekend alcohol consumption, the average grade changes between $-0.232$ and $0.685$.
*interaction*: When the gender of student is female, we are $95\%$ confident that as 1 level increases in weekend alcohol consumption, the average grade changes between $-0.232$ and $0.685$. When the gender of students is male, we are $95\%$ confident that as 1 level increases in weekend alcohol consumption, the average grade changes between $-1.720$ and $0.367$. 


### Lasso ###
```{r, include = FALSE, echo = FALSE}
X <- model.matrix(AvgGrade ~ dalc + walc + absences + fedu + medu + sex, data = Student_mat)[,-1]
y <- as.numeric(Student_mat$AvgGrade)

fit.lasso <- glmnet(X, y, alpha = 1)
plot(fit.lasso, xvar = "lambda", label = TRUE)

lasso_model <- glmnet(X, y, alpha = 1, lambda = 0.001)
coef(lasso_model, s = 0.001)

fit.lasso.cv <- cv.glmnet(X, y, alpha = 1, nfolds = 5)
plot(fit.lasso.cv)

fit.lasso.cv$lambda.min
log(fit.lasso.cv$lambda.min)
fit.lasso.cv$lambda.1se
log(fit.lasso.cv$lambda.1se)

coef(fit.lasso.cv, s = "lambda.min")
coef(fit.lasso.cv, s = "lambda.1se")
```
New predictors: father's education and workday alcohol consumption

```{r, include = FALSE, echo = FALSE}
Model9 <- lm(AvgGrade ~ dalc + walc + fedu + medu + sex, data = Student_mat)
summary(Model9)
```
Model9 is pretty close to the Model8. But there is something random in lasso. 

\begin{center}
References 
\end{center}

Riordan, Benjamin C., et al. “Just a First-Year Thing? The Relations between Drinking During Orientation Week and Subsequent Academic Year Drinking Across Class Years.” Substance Use & Misuse, vol. 53, no. 9, 2018, pp. 1501–1510., doi:10.1080/10826084.2017.1415354. 
https://web-b-ebscohost-com.ezproxy.stolaf.edu/ehost/detail/detail?vid=0&sid=792a196f-9472-4864-b717-56cc6e6b832f%40pdc-v-sessmgr03&bdata=JnNpdGU9ZWhvc3QtbGl2ZQ%3d%3d#AN=130633250&db=aph   

The summary of this article in the above link, is included in the first paragraph of the proposal. 

Graaf, Paul M. De. “Parents' Financial and Cultural Resources, Grades, and Transition to Secondary School in the Federal Republic of Germany.” European Sociological Review, vol. 4, no. 3, 1988, pp. 209–221., doi:10.1093/oxfordjournals.esr.a036485. 
https://www.jstor.org/stable/522807?seq=1#metadata_info_tab_contents  

El Ansari, Walid., et al. “Is Alcohol Consumption Associated with Poor Academic Achievement in University Students?”. International Journal of Preventive Medicine, 2013.
Is Alcohol Consumption Associated with Poor Academic Achievement in University Students? 

Research on the association between alcohol consumption and academic performance has been mixed with some  research supporting the association while others controlling and seeing labeling it as negligible. This article concluded that alcohol misuse particularly alcohol abuse is like likely to have an affect on academic performance.